Retrofitting Effect Handlers onto OCaml

### 2021-03-24 retro-concurrency.tex
minor edit
updates

We now present the implementation details of effect handlers in Multicore
OCaml. We assume an x86\_64 architecture for the remainder of this paper.
Although Multicore OCaml currently only supports the x86\_64 backend on Linux,
our design does not preclude other CPUs and OSs.

@@

Logically, DWARF call-frame information maintains a large table which records
for every machine instruction where the return address and callee-saved
registers are stored. To avoid reifying this large table, DWARF directives
represent the table using a compact bytecode representation. This representation
is interpreted on demand for building the stack unwind table. Within each
function, DWARF maintains a \emph{canonical frame address} (CFA) and is
traditionally the stack pointer before entering this function. Hence, on x86-64,
where the return address is pushed on the stack on call, the return address is
at |CFA - 8|.

The goal is to encode the CFA using DWARF directives when stacks are switched.
Recall that stack switching occurs in effect handlers, external calls and
callbacks. When starting to run on a new fiber, we insert DWARF bytecode to
follow the |parent_fiber| pointer and dereference the |saved_sp| to get the CFA
(|saved_sp + 8|). During callbacks into OCaml, we save the current system stack
pointer in the |context block| in Figure~\ref{fig:fiber} to identify the CFA in
the C stack. DWARF unwinding for external calls is implemented by following a
link to the current OCaml stack pointer. With these changes, we get the same
backtrace for the \texttt{meander} program from \S\ref{sec:unwind}, modulo
runtime system functions due to effect handlers.

@@

Despite the correct DWARF unwind information, using DWARF to record call stack
information in |perf| only captures the call stack of the current fiber in
Multicore OCaml. Since stack unwinding using DWARF is slow due to bytecode
interpretation overhead, |perf| dumps the (user) call stack when
sampled~\cite{Bastian19}. This only includes the frames from the current fiber.
This is a limitation of |perf| and not of our stack layout.

@@

Multicore OCaml uses the call stack for implementing continuations (as
do~\cite{Leijen14, Kiselyov12}), but with one-shot continuations. Bruggeman et
al.~\cite{Bruggeman96} show how to implement one-shot continuations efficiently
using segmented stacks in Scheme. Farvardin et al.~\cite{Farvardin20} perform a
comprehensive evaluation of various implementation strategies for continuations
on modern hardware. Multicore OCaml stacks do not neatly fit the description of
one of these implementation strategies -- they are best described as using the
resize strategy from Farvardin et al. for each of the fibers, which are
linked to represent the current stack and the captured continuations.

Kawahara et al.~\cite{Kawahara20} implement one-shot effect handlers using
coroutines as a macro-expressible translation, and present an embedding in Lua
and Ruby. Lua provides asymmetric coroutines~\cite{Lua} where each coroutine
uses its own stack similar to how each handled computation runs in its own
fiber in Multicore OCaml.
%Lua does not allow coroutines to include C frames, as in
%Multicore OCaml.  - avsm/kc not quite, see https://coco.luajit.org/index.html
%It is useful to note that Lua is an interpreted
%language, unlike OCaml, and DWARF unwind tables are not a concern in Lua.

>>>>

We now present the implementation details of effect handlers in Multicore
OCaml. We assume an x86\_64 architecture for the remainder of this paper.
Although Multicore OCaml currently only supports the x86\_64 backend on Linux,
our design does not preclude other architectures and OSs.

@@

Logically, DWARF call-frame information maintains a large table which records
for every machine instruction where the return address and callee-saved
registers are stored. To avoid reifying this large table, DWARF directives
represent the table using a compact bytecode representation. Each block of such
instructions usually describe the call-frame information for a single function,
where each instruction describes the table as a delta of the state at the
previous instruction. While this leads to a compact representation, computing
the return address at a given instruction requires interpreting the bytecode
from the start of the function, which may be potentially very slow. Within each
function, DWARF maintains a \emph{canonical frame address} (CFA) and is
traditionally the stack pointer before entering this function. Hence, on
x86-64, where the return address is pushed on the stack on call, the return
address is at |CFA - 8|.

Our goal is to encode the CFA using DWARF directives when stacks are switched.
Recall that stack switching occurs in effect handlers, external calls and
callbacks. When starting to run on a new fiber, we insert DWARF bytecode to
follow the |parent_fiber| pointer and dereference the |saved_sp| to get the CFA
(|saved_sp + 8|). During callbacks into OCaml, we save the current system stack
pointer in the |context block| in Figure~\ref{fig:fiber} to identify the CFA in
the C stack. DWARF unwinding for external calls is implemented by following a
link to the current OCaml stack pointer. With these changes, we get the same
backtrace for the \texttt{meander} program from \S\ref{sec:unwind}, modulo
runtime system functions due to effect handlers.

@@

Despite the correct DWARF unwind information, using DWARF to record call stack
information in |perf| only captures the call stack of the current fiber in
Multicore OCaml. Since stack unwinding using DWARF is slow due to bytecode
interpretation overhead, |perf| dumps the (user) call stack when
sampled~\cite{Bastian19}. This only includes the frames from the current fiber.
This is a limitation of |perf| and not of our stack layout. Bastian et
al.~\cite{Bastian19} report on a technique to pre-compile the unwind table to
assembly which speeds up DWARF-based unwinding by upto 25$\times$. With this
fast unwinding, |perf| would no longer needs to dump the call stack, and hence,
we can get the backtrace of the entire call stack and not just the current
fiber.

@@

Multicore OCaml uses the call stack for implementing continuations (as
do~\cite{Leijen14, Kiselyov12}), but with one-shot continuations. Bruggeman et
al.~\cite{Bruggeman96} show how to implement one-shot continuations efficiently
using segmented stacks in Scheme. Farvardin et al.~\cite{Farvardin20} perform a
comprehensive evaluation of various implementation strategies for continuations
on modern hardware. Multicore OCaml stacks do not neatly fit the description of
one of these implementation strategies -- they are best described as using the
resize strategy from Farvardin et al. for each of the fibers, which are linked
to represent the current stack and the captured continuations. Kawahara et
al.~\cite{Kawahara20} implement one-shot effect handlers using coroutines as a
macro-expressible translation, and present an embedding in Lua and Ruby. Lua
provides asymmetric coroutines~\cite{Lua} where each coroutine uses its own
stack similar to how each handled computation runs in its own fiber in
Multicore OCaml.
